Public data: domestic dogs

### 2018-01-29 cabios-template2/bigsnpr-review.tex
need changes from H

{\color{red}
In order to perform analyses of large-scale genomic data in R, we developed two R packages, bigstatsr and bigsnpr, that provide a wide-range of building blocks which are parts of standard analyses. 
R is a programming language that makes it easy to tie together existing or new functions to be used as part of large, interactive and reproducible analyses \cite[]{R2017}.
We provide a similar format as filebacked ``big.matrix'' objects that we called ``Filebacked Big Matrices (FBMs)''. Thanks to this matrix-like format, algorithms in R/C++ can be developed or adapted for large genotype data. This data format is a particularly good trade-off between easiness of use and computation efficiency, making our code both simple and fast.
}
Package bigstatsr implements many statistical tools for several types of FBMs (unsigned char, unsigned short, integer and double). This includes implementation of multivariate sparse linear models, principal component analysis, association tests, matrix operations, and numerical summaries. The statistical tools developed in bigstatsr can be used for other types of data as long as they can be represented as matrices. Package bigsnpr depends on bigstatsr, using a special type of FBM object to store the genotypes, called ``FBM.code256''. Package bigsnpr implements algorithms which are specific to the analysis of SNP arrays, such as calls to external software for processing steps, I/O (Input/Output) operations from binary PLINK files, and data analysis operations on SNP data (thinning, testing, predicting, plotting). 
We use both a real case-control genomic dataset for celiac disease and large-scale simulated data to illustrate application of the two R packages, including two association studies and the computation of Polygenic Risk Scores (PRS). We compare results from bigstatsr and bigsnpr with those obtained when using PLINK, EIGENSOFT and PRSice. We report execution times along with the code to perform major computational tasks.

@@

For preprocessing steps, PLINK is a widely-used software. For the sake of reproducibility, one could use PLINK directly from R via systems calls. For the sake of simplicity, we provide wrappers as R functions that use system calls to PLINK for the conversion and quality control steps with as input a variety of formats (e.g.\ vcf, bed/bim/fam, ped/map) and bed/bim/fam files as output (Figure~\ref{fig:qc}).
Package bigsnpr provides fast conversions between bed/bim/fam PLINK files and the ``bigSNP'' object, which contains the genotype Filebacked Big Matrix (FBM.code256), a data frame with information on samples and another data frame with information on SNPs. We also provide another function which could be used to read from tabular-like text files in order to create a genotype in the format ``FBM''. Finally, we provide two methods for reading dosage data ([SUPPMAT]).

Most modern SNP chips provide genotype data with large call-rates. For example, the Celiac data we use in this paper presents only a proportion of 0.0004 of missing values for the genotyped SNPs after quality control. Yet, most of the functions in bigstatsr and bigsnpr do not handle missing values.
So, we provide two functions for imputing missing values of genotyped SNPs. Note that we do not impute completely missing SNPs which would require the use of reference panels and could be performed via e.g.\ imputation servers for human data \cite[]{mccarthy2016reference}. The first function is a wrapper to PLINK and Beagle \cite[]{Browning2009} which takes bed files as input and return bed files without missing values, and should therefore be used before reading the data in R (Figure~\ref{fig:impute}). Beagle uses phasing before imputing, which could require several weeks of computation even when there are only a tiny amount of missing values to be imputed. The second function is a new algorithm we developed in order to have a fast imputation method without losing much of imputation accuracy. We also provide an accurate estimation of the number of errors of imputation by SNP for post quality control.
This algorithm is based on Machine Learning approaches for genetic imputation \cite[]{Wang2012} and does not use phasing, thus allowing for a dramatic decrease in computation time. It only relies on some local XGBoost models. XGBoost is an optimized distributed gradient boosting library that can be used in R and provides some of the best results in machine learning competitions \cite[]{Chen2016}.
XGBoost builds decision trees that can detect nonlinear interactions, partially reconstructing phase, making it well suited for imputing genotype matrices. 
Our algorithm is the following: for each SNP, we divide the individuals in the ones which have a missing genotype (test set) and the ones which have a non-missing genotype for this particular SNP. Those latter individuals are further separated in a training set and a validation set (e.g.\ 80\% training and 20\% validation). The training set is used to build the XGBoost model for predicting missing data. The prediction model is then evaluated on the validation set for which we know the true genotype values, which gives an unbiased estimator of the number of genotypes that have been wrongly imputed for that particular SNP. The prediction model is also projected on the test set (missing values) in order to impute them.
}

@@

SNP thinning improves ascertainment of population structure with PCA \cite[]{Abdellaoui2013}. There are at least 3 different approaches to thin SNPs based on Linkage Disequilibrium. Two of them, named pruning and clumping, address SNPs in LD close to each others because of recombination events, while the third one address long-range regions with a complex LD pattern due to other biological events such as inversions \cite[]{Price2008}. 
First, pruning is an algorithm that sequentially scan the genome for nearby SNPs in LD, performing pairwise thinning based on a given threshold of correlation.
{\color{red}
Clumping is useful if a statistic is available for sorting the SNPs by importance. Clumping is usually used to post-process results of a GWAS in order to keep only the most significant SNP per region of the genome. 
For PCA, the thinning procedure should remain unsupervised (one must not use any phenotype) and we therefore propose to use the minor allele frequency (MAF) as the statistic of importance. 
This choice is consistent with the pruning algorithm of PLINK; when two nearby SNPs are correlated, PLINK keeps only the one with the highest MAF.
Yet, in some worst-case scenario, the pruning algorithm can leave regions of the genome without any representative SNP at all ([SUPMAT]). 
So, we suggest to always use clumping instead of pruning, using the MAF as the statistic of importance, which is the default in function snp\_clumping of package bigsnpr. In practice, for the three datasets we considered, the clumping algorithm with the MAF provides similar sets of SNPs as when using the pruning algorithm (result not shown).
}

@@

Authors acknowledge Grenoble Alpes Data Institute, supported by the French National Research Agency under the ``Investissements d'avenir'' program (ANR-15-IDEX-02) and the LabEx PERSYVAL-Lab (ANR-11-LABX-0025-01). 
We would like to thank the reviewers of this paper because their comments and suggestions have led to a significant improvement of this paper.

>>>>

{\color{red}
In order to perform analyses of large-scale genomic data in R, we developed two R packages, bigstatsr and bigsnpr, that provide a wide-range of building blocks which are parts of standard analyses. 
R is a programming language that makes it easy to tie together existing or new functions to be used as part of large, interactive and reproducible analyses \cite[]{R2017}.
We provide a similar format as filebacked ``big.matrix'' objects that we called ``Filebacked Big Matrices (FBMs)''. Thanks to this matrix-like format, algorithms in R/C++ can be developed or adapted for large genotype data. This data format is a particularly good trade-off between easiness of use and computation efficiency, making our code both simple and fast.
}
Package bigstatsr implements many statistical tools for several types of FBMs (unsigned char, unsigned short, integer and double). This includes implementation of multivariate sparse linear models, principal component analysis, association tests, matrix operations, and numerical summaries. The statistical tools developed in bigstatsr can be used for other types of data as long as they can be represented as matrices. Package bigsnpr depends on bigstatsr, using a special type of FBM object to store the genotypes, called ``FBM.code256''. Package bigsnpr implements algorithms which are specific to the analysis of SNP arrays, such as calls to external software for processing steps, I/O (Input/Output) operations from binary PLINK files, and data analysis operations on SNP data (thinning, testing, predicting, plotting). 
We use both a real case-control genomic dataset for celiac disease and large-scale simulated data to illustrate application of the two R packages, including two association studies and the computation of Polygenic Risk Scores (PRS). We compare results from bigstatsr and bigsnpr with those obtained when using PLINK, EIGENSOFT and PRSice. We report execution times along with the code to perform major computational tasks. 
{\color{red}
For a comprehensive comparison between R packages bigstatsr and bigmemory, see supplementary notebook "bigstatsr-and-bigmemory".
}

@@

For preprocessing steps, PLINK is a widely-used software. For the sake of reproducibility, one could use PLINK directly from R via systems calls. For the sake of simplicity, we provide wrappers as R functions that use system calls to PLINK for the conversion and quality control steps with as input a variety of formats (e.g.\ vcf, bed/bim/fam, ped/map) and bed/bim/fam files as output (Figure~\ref{fig:qc}).
Package bigsnpr provides fast conversions between bed/bim/fam PLINK files and the ``bigSNP'' object, which contains the genotype Filebacked Big Matrix (FBM.code256), a data frame with information on samples and another data frame with information on SNPs. We also provide another function which could be used to read from tabular-like text files in order to create a genotype in the format ``FBM''. Finally, we provide two methods for reading dosage data (Supplementary notebook ``dosage'').

Most modern SNP chips provide genotype data with large call-rates. For example, the Celiac data we use in this paper presents only a proportion of 0.04\% of missing values for the genotyped SNPs after quality control. Yet, most of the functions in bigstatsr and bigsnpr do not handle missing values.
So, we provide two functions for imputing missing values of genotyped SNPs. Note that we do not impute completely missing SNPs which would require the use of reference panels and could be performed via e.g.\ imputation servers for human data \cite[]{mccarthy2016reference}. The first function is a wrapper to PLINK and Beagle \cite[]{Browning2009} which takes bed files as input and return bed files without missing values, and should therefore be used before reading the data in R (Figure~\ref{fig:impute}). Beagle uses phasing before imputing, which could require several weeks of computation even when there are only a tiny amount of missing values to be imputed. The second function is a new algorithm we developed in order to have a fast imputation method without losing much of imputation accuracy. We also provide an accurate estimation of the number of errors of imputation by SNP for post quality control.
This algorithm is based on Machine Learning approaches for genetic imputation \cite[]{Wang2012} and does not use phasing, thus allowing for a dramatic decrease in computation time. It only relies on some local XGBoost models. XGBoost is an optimized distributed gradient boosting library that can be used in R and provides some of the best results in machine learning competitions \cite[]{Chen2016}.
XGBoost builds decision trees that can detect nonlinear interactions, partially reconstructing phase, making it well suited for imputing genotype matrices. 
Our algorithm is the following: for each SNP, we divide the individuals in the ones which have a missing genotype (test set) and the ones which have a non-missing genotype for this particular SNP. Those latter individuals are further separated in a training set and a validation set (e.g.\ 80\% training and 20\% validation). The training set is used to build the XGBoost model for predicting missing data. The prediction model is then evaluated on the validation set for which we know the true genotype values, providing an estimator of the number of genotypes that have been wrongly imputed for that particular SNP. The prediction model is also projected on the test set (missing values) in order to impute them.
}

@@

SNP thinning improves ascertainment of population structure with PCA \cite[]{Abdellaoui2013}. There are at least 3 different approaches to thin SNPs based on Linkage Disequilibrium. Two of them, named pruning and clumping, address SNPs in LD close to each others because of recombination events, while the third one address long-range regions with a complex LD pattern due to other biological events such as inversions \cite[]{Price2008}. 
First, pruning is an algorithm that sequentially scan the genome for nearby SNPs in LD, performing pairwise thinning based on a given threshold of correlation.
{\color{red}
Clumping is useful if a statistic is available for sorting the SNPs by importance. Clumping is usually used to post-process results of a GWAS in order to keep only the most significant SNP per region of the genome. 
For PCA, the thinning procedure should remain unsupervised (one must not use any phenotype) and we therefore propose to use the minor allele frequency (MAF) as the statistic of importance. 
This choice is consistent with the pruning algorithm of PLINK; when two nearby SNPs are correlated, PLINK keeps only the one with the highest MAF.
Yet, in some worst-case scenario, the pruning algorithm can leave regions of the genome without any representative SNP at all (Supplementary notebook ``pruning-vs-clumping''). 
So, we suggest to always use clumping instead of pruning, using the MAF as the statistic of importance, which is the default in function snp\_clumping of package bigsnpr. In practice, for the three datasets we considered, the clumping algorithm with the MAF provides similar sets of SNPs as when using the pruning algorithm (results not shown).
}

@@

Authors acknowledge LabEx PERSYVAL-Lab (ANR-11-LABX-0025-01). Authors also acknowledge the Grenoble Alpes Data Institute that is supported by the French National Research Agency under the ``Investissements d'avenir'' program (ANR-15-IDEX-02).
We would like to thank the reviewers of this paper because their comments and suggestions have led to a significant improvement of this paper.

### 2018-01-29 public-data.Rmd
need changes from H

```{r, eval=FALSE}
bed <- "http://www.datadryad.com/bitstream/handle/10255/dryad.104774/cornell_canine.bed"
bim <- "http://www.datadryad.com/bitstream/handle/10255/dryad.104773/cornell_canine.bim"
fam <- "http://www.datadryad.com/bitstream/handle/10255/dryad.104772/cornell_canine.fam"
lapply(c(bed, bim, fam), function(url) {
  download.file(url, destfile = paste0("backingfiles/", basename(url)))
})
```

```{r}
plink <- download_plink()
snp_plinkQC("backingfiles/cornell_canine", 
            plink.path = plink,
            maf = 0.05,
            geno = 0,
            hwe = 1e-10,
            autosome.only = TRUE,
            extra.options = "--dog")
```

>>>>

```{r}
bed <- "http://www.datadryad.com/bitstream/handle/10255/dryad.104774/cornell_canine.bed"
bim <- "http://www.datadryad.com/bitstream/handle/10255/dryad.104773/cornell_canine.bim"
fam <- "http://www.datadryad.com/bitstream/handle/10255/dryad.104772/cornell_canine.fam"
lapply(c(bed, bim, fam), function(url) {
  download.file(url, destfile = paste0("backingfiles/", basename(url)))
})
```

```{r}
library(bigsnpr)
plink <- download_plink()
snp_plinkQC("backingfiles/cornell_canine", 
            plink.path = plink,
            maf = 0.05,
            geno = 0,
            hwe = 1e-10,
            autosome.only = TRUE,
            extra.options = "--dog")
```
